Parallel global optimization algorithm employing decision trees for launching local methods

### 2025-10-11 2025/MMST_DecisionTree/ParallelDecisionTree.tex
major correction

We consider the problem of finding the global minimum of the function $\varphi(y)$ over a hyperinterval $D$, 
\begin{equation} \label{sec:problem}   
	\varphi(y^*) = min\{\varphi(y):y\in D\}, D = \{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N \},
\end{equation}
where $a,b \in R$ are given vectors.

We assume that the function satisfies the Lipschitz condition 
\begin{displaymath} 
	|\varphi(y_1)-\varphi(y_2)|\leq L\parallel y_1-y_2 \parallel ,y_1,y_2 \in D, 0<L< \infty, 
\end{displaymath}
with an a priori unknown constant $L$, which corresponds to a bounded change in the function's values for a bounded change in the argument.
This assumption can be viewed (in the context of real-world problems) as reflecting the limited power generating changes in the simulated system.

The numerical solution of the problem (\ref{sec:problem}) reduces to constructing an estimate $y_k^\ast\in D$ that corresponds to some notion of proximity to a point $y^\ast$ (e.g., ${ ||y^\ast-y}_k^\ast||\le\ \varepsilon$, where $\varepsilon\geq0$ is a given accuracy) based on a finite number $k$ of computed values of the objective function. With respect to the class of problems considered, we assume that the objective function $\varphi(y)$ can be algorithmically defined through the execution of a subroutine.

When solving multidimensional problems, dimensionality reduction is employed (i.e., reducing the multidimensional problem to an equivalent one-dimensional one) using Peano curves. These curves allow transforming the multidimensional optimization problem in the region $D$ into a one-dimensional minimization problem on the interval [0, 1].
\begin{displaymath}
	\varphi(y(x^\ast))\ =\min\{\varphi(y(x)): x\in [0,1]\},
\end{displaymath}
where function $\varphi(y(x^\ast))$ satisfies the more general Hölder condition
\begin{displaymath}
	\left|\varphi (y \left(x_1\right))- \varphi (y \left(x_2\right)\right )|\le\ H\left|x_1-x_2\right|^\frac{1}{N},\ x_1,\ x_2\epsilon[0,1].
\end{displaymath}

Consequently, rather than addressing the original problem of minimizing the function $\varphi(y)$ within the domain $D$, we can consider minimizing the one-dimensional function $f(x)=\varphi(y(x))$, subject to the Hölder condition for $x \in [0, 1]$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This approach is discussed in the works \cite{Sergeyev2022, Usova2024}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Чтобы определить, в какой момент лучше запускать локальный метод в рамках глобального поиска, воспользуемся алгоритмом, основанным на дереве решений. На первом этапе поиска происходит накопление поисковой информации в виде результатов выполненных испытаний. На основе результатов испытаний проводилось обучение дерева решений, позволяющего получать кусочно-постоянную аппроксимацию целевой функции и на ее основе прогнозировать поведение целевой функции. 
To identify when to best launch a local method  within the framework of a global search, we employ an algorithm based on a decision tree  \cite{Barkalov2023_2}. The first stage of the search involves accumulating search information in the form of results from performed trials. Based on these trial results, a decision tree is trained, allowing us to obtain a piecewise-constant approximation of the objective function and, consequently, to predict the behavior of the objective function.

@@

From a formal point of view, the model used is a binary tree, that is, a tree structure where each internal (non-leaf) node has exactly two child nodes. This versatile tool can be used to solve two types of problems: classification and regression. In classification tasks, the result of processing an object is to assign it to one of the classes, for which each terminal node (leaf) of the tree is marked with the corresponding class label, while several leaves may contain identical labels. In regression problems, which is more relevant for our case, each sheet is assigned a certain constant (scalar) value. Thus, the final regression model is a piecewise constant function that takes fixed values in different parts of the definition area.

In the framework of the proposed approach, a decision tree is used to construct an auxiliary function $\psi(y)$, which serves as a piecewise constant approximation of a given function $\varphi(y)$ within the search area under study. The value calculated by this model at any point $y$ is denoted as $z' = \psi(y)$. This approximation makes it possible to effectively evaluate the behavior of the objective function and make decisions about the further direction of computing resources in the optimization process.

@@

Fig. \ref{fig:fig2} shows the graph of the objective function $\varphi(y)$; Fig. \ref{fig:fig2_2} presents the corresponding approximation $\psi(y)$ constructed using a decision tree.
\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.7\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig5.png}
			\caption{Graph of the benchmark objective function $\varphi(y)$} %% подпись к рисунку
			\label{fig:fig2}
		\end{minipage}
	\end{center}
\end{figure}

@@

The exploratory search is conducted as follows:
\begin{itemize}[label=$\bullet$] 
	\item The step size is determined (it is different for each coordinate direction and can change during the search);
	\item A search step is considered successful if the objective function value at the new point does not exceed the value at the initial point; 
	\item Otherwise, it is necessary to return and take a step in the opposite direction; 
	\item After iterating through all $N$ coordinates, the exploratory search ends; the resulting point is called the base point.
\end{itemize}
The pattern search consists of taking one step from the resulting base point along the line connecting it to the previous base point.

@@

@@

In the proposed approach, the parallelization scheme follows a ``master-worker'' principle. The master process executes the global search algorithm, accumulates search information, calculates the Lipschitz constant for the objective function based on this information, determines new trial points, and distributes them among the worker processes. The worker processes receive the trial points from the master process, evaluate the function at those points, and send the results back to the master process.

At the initial stage, the master process (let it have the identifier 0) launches $p$ trials in parallel at $p$ different points $\{y\left(x^1\right),y\left(x^2\right),\ldots,y\left(x^p\right)\}$, where preimages $\{x^1,x^2,\ldots,x^p\}$ are form the interval $[0,1]$.
Two preimages are the boundary points $x^1=0, x^p=1$; the remaining preimages are interior points $x^i\in\left(0,1\right),i=2,\ldots,p-1$.

@@

The following $p$ trials are executed in parallel at the points$x^{k+j},\ 1\leq j\leq p$,  computed using formula  (\ref{agp5_x1}). Upon completion of these trials, their results are stored in the search information database, and the algorithm proceeds to compute new trial points. Note that, generally, the trial process in applied optimization problems is far more computationally expensive than computing the coordinates of a point.

The algorithm terminates if the condition \(\Delta_{t_j} < \varepsilon\) is satisfied for at least one value of $t_j,\ 1\le\ j\le\ p$, from (\ref{agp4_R_sort} ). This stopping criterion (along with the usual criterion of limiting the number of iterations, common for iterative methods) is used in applied optimization problems where the global minimum point $y^*$ is not known a priori.

When solving test problems where the global minimizer $y^*$ is known, a stopping criterion based on reaching a neighborhood of the global minimum can be used. In this case, the method stops if the condition $\left\|y(x_{t_j})-\ y^\ast\right\| < \varepsilon$ is satisfied for one value of $t_j,\ 1\le\ j\le\ p$ from (\ref{agp4_R_sort}).

The values  
\begin{equation} 
	f_k^*=\min_{1\leq i \leq k}f(x_i), \; x_k^*=arg \min_{1\leq i \leq k}f(x_i). 
\end{equation} 
are taken as the final estimate of the global minimum for the problem under consideration.

See \cite{Strongin2000,Barkalov2016} for the justification of this computational approach. Modifications to account for inequality constraints, as well as information about the derivative of the objective function, are presented in \cite{Barkalov2002, Gergel1997, Barkalov2023, Gegrel2021}.

% переводить до этого раздела
\section{Numerical experiments}\label{SecR}

%Численные эксперименты проводились на суперкомпьютере Лобачевский. Каждый узел имел по два процессора Intel Sandy Bridge E5-2660 2,2 ГГц, 64 Гб оперативной памяти.
Numerical experiments were conducted on the Lobachevsky supercomputer. Each supercomputer node had two Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 GB of RAM.

%В экспериментах использовался генератор тестовых задач GKLS, который может генерировать задачи многоэкстремальной оптимизации с известными свойствами: точка глобального минимума, количество локальных минимумов и др.
The GKLS test problem generator was used in the experiments. This generator can create multi-extremal optimization problems with known properties, such as the location of the global minimum, the number of local minima, etc.

@@

The results below show the performance of a synchronous parallel global search algorithm using a decision tree to find basins of attraction for local minima. The numerical comparison was performed on two classes of GKLS functions (Simple and Hard from \cite{Sergeyev2006}) with dimensions 2, 3, 4, and 5. These two classes differ in the size of the basin of attraction for the global minimum. The simple class of problems has a radius of attraction that is three times larger than that of the hard class.

>>>>

The problem under consideration is to find the global minimum point of the function $\varphi(y)$ over a hyperinterval $D$, 
\begin{equation} \label{sec:problem}   
	\varphi(y^*) = \min\{\varphi(y):y\in D\}, \; D = \{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N \},
\end{equation}
where $a,b \in R$ are given vectors.

We assume that the function satisfies the Lipschitz condition 
\begin{displaymath} 
	|\varphi(y')-\varphi(y'')|\leq L \left\| y'-y'' \right\|, \; y',y'' \in D, 0<L< \infty, 
\end{displaymath}
with a constant $L$ that is unknown a priori.
This condition corresponds to a bounded change in the function's values for a bounded change in the argument.
The assumption that the objective function satisfies the Lipschitz condition can be considered (in the context of applied problems) as a reflection of the bounded possible variations in the system under optimization.

Numerically solving the problem (\ref{sec:problem}) entails constructing an estimate $y^k\in D$ aligns with a specific concept of closeness to the point $y^\ast$ (e.g., $\left\|y^\ast-y^k \right\| \le\ \delta$, where $\delta\geq0$ is a given accuracy) based on a finite number $k$ of computed function values. With respect to the class of problems considered, we assume that the function $\varphi(y)$ can be algorithmically defined through the execution of a subroutine.

The approach under consideration employs a dimensionality reduction scheme using Peano curves to solve multidimensional problems.
These curves enable the conversion of a multidimensional optimization problem within the region $D$ into a one-dimensional minimization problem over an interval $[0,1]$,
\begin{displaymath}
	\varphi(y(x^\ast))\ =\min\{\varphi(y(x)): x\in [0,1]\},
\end{displaymath}
where function $\varphi(y(x^\ast))$ satisfies the Hölder condition
\begin{displaymath}
	\left|\varphi (y \left(x_1\right))- \varphi (y \left(x_2\right)\right )|\le\ H\left|x_1-x_2\right|^\frac{1}{N},\ x_1,\ x_2\epsilon[0,1].
\end{displaymath}

Consequently, rather than addressing the original problem of minimizing the function $\varphi(y)$ within the region $D$, we can consider minimizing the one-dimensional function $f(x)=\varphi(y(x))$ for $x \in [0, 1]$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This approach is discussed in the works \cite{Sergeyev2022, Usova2024}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Чтобы определить, в какой момент лучше запускать локальный метод в рамках глобального поиска, воспользуемся алгоритмом, основанным на дереве решений. На первом этапе поиска происходит накопление поисковой информации в виде результатов выполненных испытаний. На основе результатов испытаний проводилось обучение дерева решений, позволяющего получать кусочно-постоянную аппроксимацию целевой функции и на ее основе прогнозировать поведение целевой функции. 
To identify when to best launch a local method  within the framework of a global search, we employ an algorithm based on a decision tree  \cite{Barkalov2023_2}. The first phase of the search involves accumulating search information in the form of results from performed trials.
Based on these trial results, a decision tree is trained to obtain a piecewise-constant approximation of the objective function for predicting its behavior.

@@

From a formal point of view, the model used is a binary tree, that is, a tree structure where each internal node has exactly two child nodes. This versatile tool can be used to solve two types of problems: classification and regression. In classification tasks, the result of processing an object is to assign it to one of the classes, for which each terminal node of the tree is marked with the corresponding class label, while several leaves may contain identical labels. In regression problems, which is more relevant for our case, each sheet is assigned a certain constant (scalar) value. Thus, the final regression model is a piecewise constant function that takes fixed values in different parts of the definition area.

In the framework of the proposed approach, a decision tree is used to construct an auxiliary function $\psi(y)$, which serves as a piecewise-constant approximation of a given function $\varphi(y)$ within the search area under study. The value calculated by this model at any point $y$ is denoted as $z' = \psi(y)$. This approximation makes it possible to effectively evaluate the behavior of the objective function and make decisions about the further direction of computing resources in the optimization process.

@@

Fig. \ref{fig:fig2} shows the graph of the function $\varphi(y)$; Fig. \ref{fig:fig2_2} illustrates its approximation $\psi(y)$ constructed using a decision tree.
\begin{figure}
	\begin{center}
		\begin{minipage}[h]{0.7\linewidth}
			\includegraphics[width=1\linewidth]{figure/fig5.png}
			\caption{Graph of the benchmark objective function $\varphi(y)$} %% подпись к рисунку
			\label{fig:fig2}
		\end{minipage}
	\end{center}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Исследующий поиск определяется следующим образом: 
%\begin{itemize}[label=$\bullet$] 
%  \item Вначале определяется величина шага: она является различной для каждого координатного направления и может изменяться во время поиска. 
%  \item Шаг поиска считается успешным, если значение целевой функции в новой точке не превышает значение целевой функции в исходной точке. 
%  \item Если шаг является неуспешным, то нужно вернуться к исходной точке и сделать шаг в обратном направлении. 
%  \item После перебора всех N координатных направлений исследовательский поиск завершается; полученная точка является базовой точкой для поиска по образцу.
%\end{itemize}
%Поиск по образцу состоит из выполнения одного шага от полученной базовой точки вдоль линии, соединяющей ее с предыдущей базовой точкой.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The exploratory search is conducted as follows:
\begin{itemize}[label=$\bullet$] 
	\item Initially, a step size is determined; it can be different for each coordinate direction and can change during the search;
	\item A search step is considered successful if the value of the objective function at the new point does not exceed the value of the objective function at the starting point; 
	\item  If a step is unsuccessful, it is necessary to return to the starting point and take a step in the opposite direction; 
	\item After iterating through all $N$ coordinate directions, the exploratory search is completed; the resulting point is the base point for the pattern search.
\end{itemize}

Pattern search involves moving from the current base point along the line connecting it to the previous base point.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%In the proposed approach, the parallelization scheme follows a ``master-worker'' principle. The master process executes the global search algorithm, accumulates search information, calculates the Lipschitz constant for the objective function based on this information, determines new trial points, and distributes them among the worker processes. The worker processes receive the trial points from the master process, evaluate the function at those points, and send the results back to the master process.

Our approach employs a ``master-worker'' parallelization strategy. The master process manages the global search algorithm, collects data to estimate the Lipschitz constant, computes new trial points, and assigning these points to the worker processes. The workers then evaluate the objective function at their assigned trial points and send the function values back to the master.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For the convenience of the subsequent algorithm description, we introduce the following notation.
\begin{itemize}
    \item By a \textit{trial} (or conducting a trial) we mean the calculation of the objective function value $z = f(x) = \varphi(y(x))$;
    \item By a \textit{trial point} we mean a point $y \in D$ and its preimage $x\in [0,1]$, where $y=y(x)$;
    \item We assume that the computational system used has $p + 1$ computing devices; trials will be conducted or a local method will be launched on $p$ of these devices (``workers''), while the main steps of the algorithm will be executed on one device (``master'');
    \item Let $U$ denote the set of trial points located in the basin of attraction of a minimum (local or global), from which a local method will be launched;
    \item Let $V$ denote the set of trial points which will be checked for the practicability of launching a local method from them; points that pass this check are placed into set $U$;
    \item Let $c$ is the number of points from which a local method needs to be launched at the current iteration.
    \item Let $n$ is the number of points at which trials will be conducted at the current iteration.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

At the initial stage, the master process (let it have the identifier 0) launches $p$ trials in parallel at $p$ different points $\{y\left(x^1\right),y\left(x^2\right),\ldots,y\left(x^p\right)\}$, where preimages $\{x^1,x^2,\ldots,x^p\}$ are from the interval $[0,1]$.
Two preimages are the boundary points $x^1=0, x^p=1$; the remaining preimages are interior points $x^i\in\left(0,1\right),i=2,\ldots,p-1$.
The sets $U$ and $V$ are empty.

@@

The next $p$ trials are executed in parallel at the points $x^{k+j},\ 1\leq j\leq p$,  computed using formula  (\ref{agp5_x1}). Upon completion of these trials, their results are stored in the search information database, and the algorithm proceeds to compute new trial points. Note that, generally, the trial process in applied optimization problems is far more computationally expensive than computing the coordinates of a point.

The algorithm terminates if the inequality \(\Delta_{t_j} < \delta \) is true for at least one value of $t_j,\ 1\le\ j\le\ p$, from (\ref{agp4_R_sort}).
%This stopping condition (along with the usual criterion of limiting the number of iterations, common for iterative methods) is used in applied optimization problems where the global minimizer $y^*$ is not known a priori.
In applied optimization problems, where the global minimizer $y^*$ is typically not known in advance, this stopping condition complements the standard practice of limiting iterations.

When solving test problems where the global minimizer $y^*$ is known, a stopping criterion based on reaching a neighborhood of the global minimum point can be used. In this case, the algorithm terminates if the inequality $\left\|y(x_{t_j})-\ y^\ast\right\| < \delta$ is true for one value of $t_j,\ 1\le\ j\le\ p$ from (\ref{agp4_R_sort}).

%The values  
%\begin{equation} 
%	f_k^*=\min_{1\leq i \leq k}f(x_i), \; x_k^*=arg \min_{1\leq i \leq k}f(x_i),
%\end{equation} 
%are taken as the final estimate of the global minimum for the problem under consideration.

See \cite{Strongin2000,Barkalov2016} for the justification of this computational approach. Techniques for handling inequality constraints and using objective function derivative are discussed in \cite{Barkalov2002, Gergel1997, Barkalov2023, Gegrel2021}.

% переводить до этого раздела
\section{Computational experiments}\label{SecR}

%Численные эксперименты проводились на суперкомпьютере Лобачевский. Каждый узел имел по два процессора Intel Sandy Bridge E5-2660 2,2 ГГц, 64 Гб оперативной памяти.
The Lobachevsky supercomputer was used for the numerical experiments. Each supercomputer node featured two Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 GB of RAM.
%В экспериментах использовался генератор тестовых задач GKLS, который может генерировать задачи многоэкстремальной оптимизации с известными свойствами: точка глобального минимума, количество локальных минимумов и др.
The GKLS test problem generator was used in the experiments. This generator can create multi-extremal optimization problems with known properties, such as the location of the global minimum, the number of local minima, etc.

@@

The following results demonstrate the performance of a synchronous parallel global search algorithm that uses decision trees to identify basins of attraction for local minima. The algorithm was tested on Simple and Hard classes of GKLS functions from \cite{Sergeyev2006} with dimensions $N=2,...,5$, which differ in the size of their global minimum's basin of attraction. The Simple class has a basin three times larger than the Hard class.
